---
title: "Bytecode internals: a single-instruction stack machine"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bytecode internals: a single-instruction stack machine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

[Odrzywołek (2026)](https://arxiv.org/abs/2603.21852), §4.1, observes
that EML expressions map naturally onto a **single-instruction stack
machine**: every node is the same operation. emlR implements that
machine literally — three opcodes, a constants pool, a preallocated
complex stack — for vectorised evaluation in the inner loop of
symbolic regression.

```{r setup}
library(emlR)
```

## The instruction set

Three opcodes (SPEC §4):

| Opcode | Mnemonic | Stack effect | Argument |
|:------:|----------|--------------|----------|
| 0 | `LIT` | `… → …, v` | constants-pool index |
| 1 | `VAR` | `… → …, x_j` | variable-table index |
| 2 | `EML` | `…, a, b → …, eml(a, b)` | none |

The compiler `compile_eml()` does a post-order tree walk, emitting one
opcode per node: `LIT` for a numeric/complex literal, `VAR` for a
name, `EML` for an interior `eml(l, r)` call.

## The K = 7 example

The paper's headline expression is `log(x) = eml(1, eml(eml(1, x), 1))`.
Compiling it:

```{r compile-log}
log_expr <- quote(eml(1, eml(eml(1, x), 1)))
bc <- compile_eml(log_expr)
str(bc, max.level = 1)
```

Seven instructions — matching K = 7 from the paper's Table 4 — with
opcodes:

```{r ops}
bc$ops
# 0=LIT, 1=VAR, 2=EML  →  LIT, LIT, VAR, EML, LIT, EML, EML
bc$args
# constant-pool / variable-table indices, 0 for EML
```

Reading the stream `LIT(1) LIT(1) VAR(x) EML LIT(1) EML EML`:

| Step | Op | Stack after |
|:---:|-----|---|
| 1 | `LIT(1)` | `[1]` |
| 2 | `LIT(1)` | `[1, 1]` |
| 3 | `VAR(x)` | `[1, 1, x]` |
| 4 | `EML` | `[1, eml(1, x)]` |
| 5 | `LIT(1)` | `[1, eml(1, x), 1]` |
| 6 | `EML` | `[1, eml(eml(1, x), 1)]` |
| 7 | `EML` | `[eml(1, eml(eml(1, x), 1))]` |

The constants pool is deduplicated: a tree with five `1`s holds a
single `1+0i` and points every `LIT` to index 1.

```{r consts}
bc$consts
bc$vars
bc$max_stack # = depth + 1 for right-deep trees
```

This is the exact RPN serialisation the paper writes as
`1 1 x E 1 E E` (with `E` denoting the `eml` operator after its
operands) — emlR's `eml_rpn()` reproduces the human-readable form:

```{r rpn}
eml_rpn(log_expr)
```

## The evaluator

`run_bytecode()` is one tight loop over the opcode stream, with a
preallocated complex stack:

```{r run}
xs <- seq(0.5, 5, length.out = 50)
out <- run_bytecode(bc, list(x = xs))
max(abs(Re(out) - log(xs)))
```

The stack slots are list cells of complex *vectors*, not single complex
scalars — when `vars$x` is length N, every slot carries a length-N
vector and every `EML` step is one vectorised
`exp(left) - log(right)`. R's BLAS-friendly `exp` and `log` do the
heavy lifting; the loop is lean.

## Equivalence with `eml_eval()`

The bytecode evaluator is by design semantically identical to
`eml_eval()`. The package's test suite asserts equivalence on 100
random EML trees of depths 1–5 over random complex input vectors:

```{r equiv}
random_eml_tree <- function(max_depth, var_names = c("x", "y")) {
  if (max_depth == 0 || runif(1) < 0.4) {
    if (runif(1) < 0.5) {
      sample(c(1, 1, 1, 0.5, 2), 1)
    } else {
      as.name(sample(var_names, 1))
    }
  } else {
    call(
      "eml",
      random_eml_tree(max_depth - 1, var_names),
      random_eml_tree(max_depth - 1, var_names)
    )
  }
}

set.seed(2026)
agree <- TRUE
for (i in seq_len(20)) {
  expr <- random_eml_tree(sample.int(5, 1))
  vars <- list(x = runif(50, 0.1, 5), y = runif(50, 0.1, 5))
  v1 <- eml_eval(expr, vars)
  v2 <- run_bytecode(compile_eml(expr), vars)
  if (!isTRUE(all.equal(v1, v2, tolerance = 1e-10))) {
    agree <- FALSE
    break
  }
}
agree
```

## When to use which evaluator

* **`eml_eval()`** — for one-off evaluation, debugging, or working with
  expressions that may not be pure-EML (`simplify_native()` output,
  for example, contains base-R primitives).
* **`compile_eml()` + `run_bytecode()`** — for the inner loop of an
  optimisation, batch evaluation over many input points, or any
  setting where the same EML expression is evaluated many times.
  Compile once, evaluate many times.

`eml_fit()` uses bytecode internally for its forward-pass evaluation;
gradients come from `Deriv::Deriv` on the `eml`-expanded master
expression.

## References

Odrzywołek, A. (2026). *All elementary functions from a single binary
operator.* [arXiv:2603.21852](https://arxiv.org/abs/2603.21852), §4.1
("EML compiler").
