---
title: "Introduction to emlR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to emlR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## The EML operator

Odrzywołek (2026, [arXiv:2603.21852](https://arxiv.org/abs/2603.21852))
proves that the single binary operator

$$\mathrm{eml}(x, y) \;=\; \exp(x) - \ln(y)$$

together with the constant `1`, generates **every elementary function**
on the principal branch of the complex numbers. emlR implements the
operator, a catalog of canonical EML expressions for every elementary
primitive, a two-mode simplifier, a stack-machine bytecode evaluator,
and an L-BFGS-B fitter for symbolic regression.

```{r setup}
library(emlR)
```

The operator itself:

```{r eml-basic}
eml(1, 1)            # exp(1) - log(1) = e
eml(0, 1)            # exp(0) - log(1) = 1
eml(0, -1)           # 1 - i*pi (principal-branch log of -1)
```

It is internally complex (the principal branch of `log` is required for
`log(-1) = iπ`). For real-valued inputs known to produce real results
use the wrapper:

```{r eml-real}
eml_real(1, 1)
eml_real(0.5, 1)
```

## Three sacred identities

The paper proves three identities that every EML simplifier must
preserve. emlR verifies them in `simplify_native()`, the rewriter that
collapses recognised EML patterns to base-R primitives.

```{r sacred}
simplify_native(quote(eml(x, 1)))                          # exp(x)
simplify_native(quote(eml(1, 1)))                          # e (folded)
simplify_native(quote(eml(1, eml(eml(1, x), 1))))           # log(x), paper Eq. 5, K=7
```

The third identity is the headline of the paper: with K = 7 nodes (or
seven RPN tokens, `1 1 x E 1 E E`), `eml(1, eml(eml(1, x), 1))` is
exactly `log(x)`.

```{r rpn}
log_expr <- quote(eml(1, eml(eml(1, x), 1)))
eml_K(log_expr)        # 7  (paper K)
eml_depth(log_expr)    # 3
eml_rpn(log_expr)      # "1 1 x E 1 E E"
```

## Native R AST

EML expressions are not a custom S3 class; they are ordinary R `call`
objects. That means every existing R language tool just works:

```{r ast}
e <- quote(eml(1, eml(eml(1, x), 1)))
deparse(e)
all.vars(e)
identical(e, bquote(eml(1, eml(eml(1, .(quote(x))), 1))))
```

You can also build expressions programmatically:

```{r constructors}
eml_node(1, "x")                   # eml(1, x)
E(1, E(E(1, "x"), 1))              # E() shorthand — paper Eq. 5
```

## Evaluating expressions

Two evaluators are available:

* `eml_eval()` — defers to `base::eval()` after binding `eml` plus your
  variables in a fresh environment.
* `compile_eml()` + `run_bytecode()` — a stack-machine fast path with a
  deduplicated constants pool. Vectorised over input bindings.

```{r eval}
xs <- seq(0.5, 5, length.out = 30)
log_expr <- quote(eml(1, eml(eml(1, x), 1)))

# Direct evaluation
v1 <- Re(eml_eval(log_expr, list(x = xs)))

# Bytecode
bc <- compile_eml(log_expr)
v2 <- Re(run_bytecode(bc, list(x = xs)))

max(abs(v1 - v2))
max(abs(v1 - log(xs)))
```

The bytecode evaluator is what powers `eml_fit()` (see
`vignette("symbolic-regression")`); it lets the inner loop of an
optimisation run vectorised over hundreds of input points without the
overhead of recursive `eval()`.

## The catalog

`eml_catalog()` returns a named list with every elementary primitive as
an EML expression:

```{r catalog}
catalog <- eml_catalog()
names(catalog)
catalog$log
catalog$sub
```

Each entry numerically evaluates to the right value (`verify_catalog()`
checks all 18 entries):

```{r verify}
verify_catalog(verbose = FALSE)
```

`simplify_native()` collapses every catalog entry — including the
trigonometric and root constructions — to its base-R form literally:

```{r catalog-simplify}
simplify_native(catalog$mul)                   # x * y
simplify_native(catalog$div)                   # x / y
simplify_native(catalog$pow)                   # x ^ y
simplify_native(catalog$sqrt)                  # x^0.5
simplify_native(catalog$sin)                   # sin(x)
simplify_native(catalog$cos)                   # cos(x)
simplify_native(catalog$pi)                    # ≈ pi
```

Reaching `sin(x)` and `cos(x)` requires the cleanup rule
`exp(_C + log(_x)) → exp(_C) * _x` to absorb numeric constants that
emerge mid-collapse, plus a small (`1e-12`) tolerance in the matcher
to absorb the round-off in `tree_i() = exp(log(-1)/2)`. Both are
applied automatically; see `?simplify_native` and SPEC §5.3 for the
full mechanism.

## Where to next

* `vignette("symbolic-regression")` — recover `log(x)` from data using
  the master formula and L-BFGS-B with analytic gradients.
* `vignette("bytecode")` — internals of the compiler and stack-machine
  evaluator.
* `?eml`, `?simplify_native`, `?eml_fit` — function reference.

## References

Odrzywołek, A. (2026). *All elementary functions from a single binary
operator.* [arXiv:2603.21852](https://arxiv.org/abs/2603.21852).
