---
title: "Symbolic regression: recovering log(x) from data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Symbolic regression: recovering log(x) from data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

This vignette reproduces the headline demo from
[Odrzywołek (2026)](https://arxiv.org/abs/2603.21852), §4.3:
recovering `ln(x)` from `(xs, ys)` data alone, with no human input
about what the closed form should be — just the EML grammar and a
gradient-based optimiser.

```{r setup}
library(emlR)
```

## The master formula

The paper's symbolic-regression construction parameterises an EML tree
of fixed depth such that each input slot is a continuous mixture of
three sources: the constant `1`, the input variable `x`, and the value
`f` flowing up from the deeper EML node. Snapping each slot's mixture
to a one-hot vertex recovers a discrete grammar choice; training the
mixture as a continuous function with gradient-based optimisation lets
the optimiser *find* the right discrete structure.

`build_master(depth)` produces the parameterised expression:

```{r master}
m1 <- build_master(1)
m1
master_n_params(1) # 4 = 5 * 2^1 - 6
```

For depth 1 there are 4 parameters: `(alpha_1, beta_1)` for the left
input and `(alpha_2, beta_2)` for the right input. Each slot is

```
alpha_i * 1 + beta_i * x        (leaf slot, no f available)
alpha_i * 1 + beta_i * x + gamma_i * f   (inner slot, f from deeper eml)
```

so a one-hot `(0, 1)` selects `x`, a one-hot `(1, 0)` selects `1`,
and `(0, 0, 1)` (inner) routes through the deeper EML node.

## A sanity-check selector

The paper's three sacred identities give us closed-form selectors. For
`exp(x) = eml(x, 1)`:

```{r exp-selector}
par_exp <- theta_for_exp()
par_exp # length 4: 0 1 1 0

xs <- seq(-1, 1, length.out = 5)
bindings <- unpack_master_params(par_exp, depth = 1)
env <- list2env(c(list(eml = eml), bindings, list(x = xs)),
  parent = baseenv()
)
Re(eval(m1, env)) # exp(xs)
```

For `log(x)` the paper's Eq. 5 expands as
`eml(1, eml(eml(1, x), 1))`, which is a depth-3 master with 34
parameters. `theta_for_log()` returns the parameter vector that
recovers it:

```{r log-selector}
m3 <- build_master(3)
master_n_params(3) # 34

par_log <- theta_for_log()
xs <- c(0.5, 1, 2, 5, 10)
bindings <- unpack_master_params(par_log, depth = 3)
env <- list2env(c(list(eml = eml), bindings, list(x = xs)),
  parent = baseenv()
)
suppressWarnings(Re(eval(m3, env)))
log(xs)
```

A perfect match, modulo machine epsilon.

## Symbolic regression: discover the structure

Now the interesting case: forget the selector and recover `log(x)`
from data alone. We give `eml_fit()` 30 noiseless `(x, log(x))` points,
parameterise the depth-3 master via per-slot softmax, and run L-BFGS-B
with analytic gradients via `Deriv::Deriv`.

The vignette deliberately uses a small budget (`n_restarts = 2`,
`maxit = 200`) so the rendered build stays under CRAN's per-vignette
time limit. The full reproducibility budget — `n_restarts = 10`,
`maxit = 1000` — is used by the Phase-5 acceptance gate at
`tests/testthat/test-master.R` (see `?eml_fit`); set
`EMLR_RUN_SLOW=true` and run that test to reproduce the paper's
~94% recovery rate.

```{r fit}
set.seed(42)
xs <- seq(0.5, 5, length.out = 30)
ys <- log(xs)

fit <- eml_fit(xs, ys,
  depth            = 3,
  parameterization = "simplex",
  method           = "L-BFGS-B",
  maxit            = 200,
  n_restarts       = 2,
  seed             = 1
)
```

`fit$snap_mse` is the MSE *after* snapping each slot to its argmax
one-hot vertex. A value at machine epsilon squared (~ `1e-32`)
indicates exact symbolic recovery — the optimiser has snapped to the
right discrete grammar.

```{r snap-mse}
fit$snap_mse
fit$best_seed # which restart won
```

Compare predicted vs. true:

```{r plot}
plot(xs, ys,
  pch = 19, ylab = "y", xlab = "x",
  main = "Recovering log(x) from data alone"
)
lines(xs, fit$pred, col = "tomato", lwd = 2)
lines(xs, fit$pred_snap, col = "darkgreen", lwd = 2, lty = 2)
legend("bottomright",
  c("data (= log(x))", "continuous fit", "snapped (one-hot)"),
  col = c("black", "tomato", "darkgreen"),
  lwd = c(NA, 2, 2), lty = c(NA, 1, 2),
  pch = c(19, NA, NA)
)
```

## Why it works

The paper's depth-3 master at the snapped optimum is *literally* the
expression `eml(1, eml(eml(1, x), 1))`. emlR's evaluator evaluates that
expression at every parameter substitution; L-BFGS-B sees a smooth loss
surface (because the softmax is smooth); analytic gradients via `Deriv`
keep the inner loop fast. With a single restart the success rate is
roughly the paper's reported ~25%; with 10 restarts the probability of
recovery exceeds 90%.

The Phase-5 acceptance gate of this package
([`tests/testthat/test-master.R`](https://github.com/pjt222/emlR/blob/main/tests/testthat/test-master.R))
asserts ≥ 8/10 multi-restart trials snap to MSE < 1e-20 on this
problem.

## Going deeper

* Use `parameterization = "direct"` to bypass softmax and fit raw
  scaling — useful for problems where you do *not* expect a discrete
  one-hot snap.
* Increase `depth` to search a larger expression space. Master formula
  parameter count is `5 * 2^depth - 6`, so depth 4 has 74 parameters.
* The expression space at depth 3 includes every K ≤ 7 EML tree, which
  is enough for `exp`, `log`, `e`, and several derived forms. Higher
  depths reach the longer constructions in the paper's Table 4.

## What this fitter is, and is not

`eml_fit()` implements the paper's master-formula SR approach: every
internal node is `eml`, every leaf is a softmax over `{x, y, 1}`, and
depth is fixed up front. Only the leaf weights are optimised; tree
*topology* is fixed by `depth`. Genetic-programming-style topology
search across the EML grammar is out of scope for this release — the
bytecode evaluator and catalog provide the building blocks if you want
to layer your own search loop on top.

## References

Odrzywołek, A. (2026). *All elementary functions from a single binary
operator.* [arXiv:2603.21852](https://arxiv.org/abs/2603.21852),
§4.3 ("Symbolic regression").
